Urbanization affects web abundance and aggregation of a funnel-weaving spider, *Agelenopsis pennsylvanica*

### 2023-01-14 manuscript/manuscript.Rmd
These are the versions that I submitted

## Keywords
Arachnids, Arthropods, Spatial Distribution, Environmental Predictors

@@

We thank Morgan Ryan for his advice about processing the GIS data in QGIS. We are grateful for past and present members of the Hebets, Shizuka, and Keesey labs (UNL Behavior Group) for their feedback on data analysis and earlier drafts of the manuscript. We also thank conference attendees that have visited posters and oral talks about this project, as those conversations have helped to improve many aspects of the research presented here.

@@

Aggregations can form when suitable conditions and resources are not evenly distributed, leading to higher-quality patches [@churchill_2020]. For example, gaps in the tree canopy allow more sunlight to reach the understory -- thus, increasing plant height and soil moisture and decreasing leaf litter -- causing some invertebrate families to become more abundant in these patches [@perry_2018]. Clumping can also result from incomplete dispersal that occurs when barriers, often due to habitat structure, prevent adequate dispersal [@riechert_1974b]. Scientists have found differences in the relative importance of different spatial predictors based on the type of taxon-specific spatial information collected -- *i.e.* presence, abundance, aggregation, or vertical position [@boulangeat_2012].

Due to intense habitat disturbance and dense human presence, urbanization has drastically modified the land [@alberti_2020; @mcdonnell_pickett_1990; @pickett_2001]. These land changes lead to altered environmental conditions, community dynamics, and dispersal capability as a result [@piano_2020]. Urban environments are characterized by decreased biodiversity [@mckinney_2002], increased impervious cover [compacted soil from buildings and pavement, @nowak_greenfield_2020], habitat patchiness [@cadenasso_2007], pollution [air, water, and light, @grimm_2008; @hopkins_2018], and noise [air-, water-, and substrate-borne, @barber_2010; @slabbekoorn_2010; @wu_elias_2014]. These conditions can contribute to changes in local climatic conditions, such as when impervious cover traps heat and increases the average temperatures in cities compared to surrounding areas --- *i.e.*, the Urban Heat Island Effect [@morabito_2021]. There is growing evidence that urbanization changes species interactions, even reshaping whole food webs [@start_2020] and altering the landscape of parasite and disease transmission [@bradley_altizer_2007]. As many of these conditions and dynamics determine animal spatial patterns, it is likely that urbanized landscapes have changed how animals spatially distribute themselves.

Urban ecosystems can also be extremely heterogeneous with unique and diverse habitats nested within [@rivkin_2019]. For example, habitats within urban ecosystems can differ considerably in landscape features (e.g. tree and building cover), weather conditions (e.g. temperature, humidity, and wind speed), and disturbance, resulting in variable access and distributions of resources and opportunities for certain species to thrive. These factors can also differentially drive animal spatial patterns [@murray_2019; @trubl_2012], sometimes leading to higher degrees of aggregation for urban animals [@johnson_munshisouth_2017].

Variation in the abundance and distribution of urban arthropods can have implications for ecosystem function due to the essential ecosystem services that arthropods provide, like pest control [@otoshi_2015] and pollination [@ivers_2022]. Arthropods are considered critical bioindicators of ecosystem health as their short generation time requires them to respond rapidly to human-induced change [@mcintyre_2000]. However, the impact of urbanization on arthropod ecology is still understudied [but see @kotze_2011; @fenoglio_2020], and the relationship between urban habitats and arthropod abundance and diversity remain controversial [@gippet_2017; @dale_frank_2018]. Such discrepancies are most likely due to abundance and distributions being species- [@gippet_2017], habitat- [@bennett_lovell_2014], and scale-dependent [@philpott_2014]. While some animals benefit from urban conditions [@guenard_2015], many cannot survive the disturbance and change associated with cities [@wagner_2020]. Thus, it is important to determine how urban factors contribute to variation in arthropod abundance and spatial patterns for the conservation of arthropods and their essential services.

@@

To understand habitat characteristics that might influence the distribution of *A pennsylvanica* across urban environments, we compared *A. pennsylvanica* spatial patterns between two urban habitats (an urban center and an urban forest). In particular, we asked the following three questions: (*i*) How do environmental conditions vary between two urban habitats? (*ii*) How do *A. pennsylvanica* web abundance, aggregation, and position vary between these two urban habitats? (*iii*) What environmental variables help predict spider spatial patterns in distinct urban habitats? Based on preliminary observations, we predicted that *A. pennsylvanica* would be more abundant and aggregated in the urban center than the urban forest. We further expected that specific environmental conditions that vary between the two habitats (e.g. tree cover, plant species richness, and disturbance from roads) would be correlated with the variation in the spatial patterns of *A. pennsylvanica*.

@@

From each start site, we began walking along a path in a randomly selected direction (North, East, South, or West using a random number generator). We conducted a thorough visual search of the edges of the path for *A. pennsylvanica* funnel webs as the GPS receiver tracked our movement (Fig. S4). We walked along pre-existing sidewalks, roads, and trails because paths were a common feature of both habitats along which we had previously found funnel-weaving spiders. We often found *A. pennsylvanica* in edge habitats in the study region, and this tendency to build webs at a habitat's edge is shared by other spider species [@cobbold_supp_2012]. Additionally, the thick undergrowth of the urban forest often restricted searches off the trails. To ensure equivalent search efforts, we took more time searching the edge of dense undergrowth in the urban forest.

@@

### Statistical Analysis

@@

Since we found significant variation in distribution measures between habitats (see Results: Urban habitat characteristics), we wanted to explore environmental correlations with our findings. Specifically, we used the habitat subsets and the selected (non-correlated) environmental predictor variables to assess relationships between predictor variables and spatial pattern measures. In the global models, we included the predictors (without interactions) that remained after removal for collinearity with the response variables: web and spider abundance, nearest neighbor distance, and web height for the complete data and habitat subsets. We restricted these nearest neighbor analyses to first nearest neighbor distance (*i.e.* excluded second nearest neighbor distance) because the first and second nearest neighbor distance both significantly differed between habitats, and including both would complicate the analysis. We excluded search distance as a response variable because the predictors were collected from the site of the focal web and may vary along the search path.

Several predictors (plant species richness, patch area, and road and highway disturbance) were highly right-skewed, so we used a natural log-transformation prior to model selection. We used Poisson generalized linear models for number of webs and spiders and nearest neighbor distance and Poisson generalized linear mixed-effect models for web height with site as a random effect. From these global models, we performed Akaike Information Criterion (AIC) model selection with the *MuMIn* package in R [@MuMIn], which ranks models produced from every combination of the predictors in the global model. We reported the results of each top model, as well as a model average of models within 2 delta AICc points of the top model. We reported the estimates and standard error from the models without back-transformations for the log-transformed variables. We checked that the Variance of Inflation (VIF) stayed at or below a score of two for predictors in each top model using the *car* package in R [@car].

We also explored how land cover may affect web abundance, aggregation, and position using the National Land Cover Database and QGIS to record the category of the land cover class for the pixel that each site was located. We tested Poisson generalized linear models for each response variable with land cover as the only predictor. We used the *Anova()* function in the *car* package [@car] to get the Likelihood Ratio Chi-Squared values, then used the *glht()* funtion in the *multcomp* package [@multcomp] to perform pairwise comparisons among categories. We include these results in supplemental material (Fig. S10, Table S2) as the sample sizes were small (deciduous forest = 7, woody wetlands = 3, urban high-intensity = 4, urban medium-intensity = 6, urban low-intensity = 2).

@@

**Table 1** Landscape variables for the urban center (UNL City Campus), the urban forest (Wilderness Park), and Lincoln (Nebraska, USA) City Limits,  collected from the National Land Cover Database (2016 Tree Cover, 2019 Impervious Cover, 2019 Land Cover), the Landsat-7 World Reference System-2 (Band 6, path 28, row 32, taken September 2, 2020), and the 2020 Visible Infrared Imaging Radiometer Suite (VIIRS) using QGIS (v. 3.16.3-Hannover, ESRI 102704)

@@

All environmental predictors assessed within a 100-meter radius of the focal web varied by habitat (Table 2). Site tree cover was lower in the urban center than the urban forest with five urban center sites measuring zero tree cover ($z_21$ = -14.50, *P* \< 0.001). The urban center sites had higher impervious cover, with all but four sites being zero in the urban forest ($z_21$ = 14.51, *P* \< 0.001). The urban center sites also had higher spectral radiance ($z_21$ = 3.08, *P* = 0.002) and artificial light at night radiance ($z_21$ = 22.46, *P* \< 0.001). Overall plant species richness was lower in the urban center than the urban forest ($z_21$ = -7.67, *P* \< 0.001). Similarly, spiders used a lower diversity of plants as substrates in the urban center (mean ± se = 1.42 ± 0.15 species) than the urban forest (mean ± se = 2.7 ± 0.26 species, $z_21$ = -2.08, *P* = 0.037). Site patch area was smaller in the urban center ($z_21$ = -374.2, *P* \< 0.001). Road disturbance ($z_21$ = 39.19, *P* \< 0.001), highway disturbance ($z_21$ = 9.39, *P* \< 0.001), and road length ($z_21$ = -374.2, *P* \< 0.001) were higher for urban center sites than sites in the urban forest.

**Table 2** Mean and standard error of environmental predictors collected in 100-meter buffers of each focal web by habitat

@@

**Fig. 3** Comparison of search distance (a), the number of webs (b) and spiders (c) in a 10-meter radius of the focal web, the distance from the focal web to the first and second (d) nearest neighbors, and web height (e) by habitat - urban center vs urban forest. Grey jittered points represent the raw data while black points and error bars represent the mean and 95% confidence intervals, respectively, from model predictions. Models were either Poisson Generalized Linear Models (a-c) or Poisson Generalized Linear Mixed Effect Models with site as a random factor (d-f). Asterisks indicate significance level (\*\*\* *P* \< 0.001; \*\* *P* \< 0.01; \* *P* \< 0.05). We searched shorter distances from the start site to the focal web (a), found more webs (b) and spiders (c) in a 10-meter radius of the focal web, measured shorter distances between the focal web and the two nearest neighbors (d), and recorded shorter web heights (e) in the urban center than the urban forest

@@

*Abundance* -- The top models for the number of webs and the number of spiders yielded similar results (Table 3, Fig. 4). For both measures of abundance, the urban forest subset retained no predictors (Table 3) while the urban center showed a decrease in abundance with increasing plant species richness, patch area, highway disturbance, and road length (Table 3, Fig. 4). The model average for the number of webs subset by urban center suggested similar results (Table S3). Plant species richness and highway disturbance both had a negative relationship with abundance measures for the overall data set (Table 3, Fig. S12). Interestingly, the top model of web abundance retained road disturbance and suggested a positive relationship overall (Table 3, Fig. S12).

@@

*Aggregation* -- Spiders were closer together across habitats when plant species richness was low and when road disturbance was high (Table 3, Fig. S12). In the urban forest subset, less tree cover, lower road and highway disturbance, and greater plant species richness were associated with nearness (Table 3, Fig. 5), however, this model was a poor fit ($R^2$ = 0.037). Urban center spiders were more aggregated where plant species richness, spectral radiance, and road length are high (Table 3, Fig. 5). The relationship between nearest neighbor distance and plant species richness is opposite when analyzed across habitats and within the urban subset. This is likely do to the differences in plant species richness and aggregation between habitats -- the urban center had low neighbor distance and plant diversity while the urban forest had high neighbor distance and plant diversity. The model average for the nearest neighbor distance in the urban center corroborates the findings of the top model and adds that webs were clumped when patch area was larger (Table S3).

@@

For statistical analysis and figures of response variables by land cover class, see supplemental material (Table S2, Fig. S10). Following Likelihood Ratio tests, we find that all response variables significantly vary by land cover class (Table S2). There was no difference in search distance between urban medium-intensity, urban low-intensity, and deciduous forest. However, we searched significantly shorter in urban high-intensity and significantly further in woody wetlands (Fig. S10a). The number of webs and spiders only varied by habitat - urban low-, medium-, and high-intensity had more webs and spiders than deciduous forest or woody wetlands (Fig. S10b-c). For the urban intensities, web distance was shortest at low intensities and furthest at medium intensities with high urban intensity suggest web height between low and medium intensities (Fig. S10d). Nearest neighbor distance was shorter for woody wetlands than deciduous forest (Fig. S10d). Lastly, web height showed no difference between land cover class with pairwise comparisons (Fig. S10e).

@@

We also found that the webs of urban center spiders were built lower than webs in the urban forest, and this result was independent of the height of the substrate. One possible explanation for this pattern is that vibratory noise promotes lower web heights. Vibratory noise might, for example, be dampened on a plant nearer to the ground. We did find a trend that showed lower web heights overall at sites with higher road disturbance, but this idea requires further testing. Another potential explanation is that webs may vary in height based on the type of prey being ensnared. We have observed *Agelenopsis* spiders capture flying, hopping, and crawling insects. Arthropod community structure can be highly variable between urban habitats [@bang_faeth_2011], and urban centers and urban forests may support different types of prey items for *A. pennsylvanica*. Lower webs in the urban center may be ideal for more crawling prey while higher webs in the urban forest may intercept more flying insects. Alternatively, webs in different habitats may experience different levels of disturbance. In the urban center where lawns are mowed frequently and ornamental plants are regularly pruned, spiders could maintain lower webs to conserve energy from frequent rebuilding. Additionally, urban forests likely support (and act as a refuge for) many ground dwelling vertebrates, whose movement along the forest floor could frequently damage lower webs. Greater tree coverage in the urban forest predicted higher webs, which may further suggest that small mammal movement could destroy low webs as areas with greater tree cover may provide protection and food resources for small mammals [@sarkar_2022]. Future research should assess the prey types available and small animal abundance and movement between these two habitats.

@@

We found that webs and spiders were more abundant (in the urban center) and more aggregated as plant species richness decreased. The urban center sites were covered in grass lawns and managed landscaping that usually had few plant species present, but in high abundance. Sites with lower plant diversity may have few plant species that are highly preferred, leading to higher spider abundance. Research has shown mixed but often positive effects of plant diversity on spider abundance [@malumbres_et_al_2013; @koricheva_et_al_2000], suggesting that plant community composition may play a vital, yet context-specific role in spider spatial patterns. To better understand the role that plant species play in abundance would require comparing plant communities between patches used by spiders versus those devoid of spiders. We speculate that greater aggregation on more diverse plant communities may suggest that *A. pennsylvanica* prefers a certain plant or type of plant, especially in the urban center where we observed similar plants clumped in space.

@@

Lower road and highway disturbance was associated with larger spider aggregations in the urban forest, but had no effect on urban center spiders. Perhaps noise negatively affects urban forest spiders -- encouraging spiders to inhabit areas further from noise sources and reducing space between neighbors. Many spiders are thought to be plastic in the structure and tuning of their webs [@mortimer_2016; @blamires_2010], and urban exploiters are often fast to adjust to changing conditions [@kralj_2017]. As such, spiders reared in or descended from spiders in the urban center might be better able to tune their webs to decrease vibrations, although this hypothesis requires further testing. It will also be important to differentiate whether these differences in behavior between environments are based on developmental exposure, behavioral plasticity, or genetic differentiation.

@@

Lastly, spiders in the urban center were more aggregated at sites with higher spectral radiance (a measure of surface temperature). Urban centers tend to be warmer on average compared to surrounding areas due to heat-trapping via pavement in a phenomenon called the Urban Heat Island Effect [@johnson_et_al_2020]. Aggregation formation in warmer sites could be the result of preference for or acclimation to warmer areas [@meineke_et_al_2013] or perhaps aggregation occurs at limited thermal refuges [@riechert_1974b].

@@

We have shown that human-caused environmental changes correlate with spider spatial patterns, but there are likely many other forms of human disturbance than those tested here. Our study shows that a solitary, generalist predator (*A. pennsylvanica*) is more abundant and aggregated with lower web height in an urban center compared to an urban forest -- habitats that vary significantly in environmental conditions. We did not measure community-level interactions among organisms, such as the effects of prey abundance; however, correlations between spatial pattern and environmental conditions indirectly indicate that such interspecific interactions may be important influencers of spider aggregations. Noise may also greatly influence spider spatial patterns. Our variables that estimate the potential for vibratory noise negatively affected spider abundance while promoting lower web heights, potentially to reduce interference from vibratory noise. Noise variables also had opposite effects on aggregative behavior of spiders in the different habitats. These differences may indicate variability in tolerance to noise, but more research is necessary to test this hypothesis.

>>>>

## Keywords

Arthropods, Environmental Predictors, Human-Induced Rapid Environmental Change (HIREC), Nearest Neighbor, Spatial Distribution

@@

We thank Morgan Ryan for his advice about processing the GIS data in QGIS. We are grateful for past and present members of the Hebets, Shizuka, and Keesey labs (UNL Behavior Group) for their feedback on data analysis and earlier drafts of the manuscript. We are also thankful for conversations with Chad Johnson about urban ecology. We thank conference attendees that have visited posters and oral talks about this project, as those conversations have helped to improve many aspects of the research presented here.

@@

Aggregations can form when suitable conditions and resources are not evenly distributed, leading to higher-quality patches [@churchill_2020]. For example, gaps in the tree canopy allow more sunlight to reach the understory -- thus, increasing plant height and soil moisture and decreasing leaf litter -- causing some invertebrate families to become more abundant in these patches [@perry_2018]. Clumping can also result from incomplete dispersal that occurs when barriers, often due to habitat structure, prevent adequate dispersal [@riechert_1974b]. Scientists have found differences in the relative importance of different spatial predictors based on the type of taxon-specific spatial information collected -- *i.e.,* presence, abundance, aggregation, or vertical position [@boulangeat_2012].

Due to intense habitat disturbance and dense human presence, urbanization has drastically modified the land [@alberti_2020; @mcdonnell_pickett_1990; @pickett_2001]. These land changes lead to altered environmental conditions, community dynamics, and dispersal capability as a result [@piano_2020]. Urban environments are characterized by decreased biodiversity [@mckinney_2002], increased impervious cover [compacted soil from buildings and pavement, @nowak_greenfield_2020], habitat patchiness [@cadenasso_2007], pollution [air, water, and light, @grimm_2008; @hopkins_2018], and noise [air-, water-, and substrate-borne, @barber_2010; @slabbekoorn_2010; @wu_elias_2014]. These conditions can contribute to changes in local climatic conditions, such as when impervious cover traps heat and increases the average temperatures in cities compared to surrounding areas --- *i.e.,* the Urban Heat Island Effect [@morabito_2021]. There is growing evidence that urbanization changes species interactions, even reshaping whole food webs [@start_2020] and altering the landscape of parasite and disease transmission [@bradley_altizer_2007]. As many of these conditions and dynamics determine animal spatial patterns, it is likely that urbanized landscapes have changed how animals spatially distribute themselves.

Urban ecosystems can also be extremely heterogeneous, with unique and diverse habitats nested within [@rivkin_2019]. For example, habitats within urban ecosystems can differ considerably in landscape features (e.g., tree and building cover), weather conditions (e.g., temperature, humidity, and wind speed), and disturbance, resulting in variable access and distributions of resources and opportunities for certain species to thrive. These factors can also differentially drive animal spatial patterns [@murray_2019; @trubl_2012], sometimes leading to higher degrees of aggregation for urban animals [@johnson_munshisouth_2017].

Variation in the abundance and distribution of urban arthropods can have implications for ecosystem function due to the essential ecosystem services that arthropods provide, like pest control [@otoshi_2015] and pollination [@ivers_2022]. Arthropods are considered critical bioindicators of ecosystem health as their short generation time requires them to respond rapidly to human-induced change [@mcintyre_2000]. However, the impact of urbanization on arthropod ecology is still understudied [but see @kotze_2011; @fenoglio_2020], and the relationship between urban habitats and arthropod abundance and diversity remains controversial [@gippet_2017; @dale_frank_2018]. Such discrepancies are most likely due to abundance and distributions being species- [@gippet_2017], habitat- [@bennett_lovell_2014], and scale-dependent [@philpott_2014]. While some animals benefit from urban conditions [@guenard_2015], many cannot survive the disturbance and change associated with cities [@wagner_2020]. Thus, it is important to determine how urban factors contribute to variation in arthropod abundance and spatial patterns for the conservation of arthropods and their essential services.

@@

To understand habitat characteristics that might influence the distribution of *A pennsylvanica* across urban environments, we compared *A. pennsylvanica* spatial patterns between two urban habitats (an urban center and an urban forest). In particular, we asked the following three questions: (*i*) How do environmental conditions vary between two urban habitats? (*ii*) How do *A. pennsylvanica* web abundance, aggregation, and position vary between these two urban habitats? (*iii*) What environmental variables help predict spider spatial patterns in distinct urban habitats? Based on preliminary observations, we predicted that *A. pennsylvanica* would be more abundant and aggregated in the urban center than the urban forest. We further expected that specific environmental conditions that vary between the two habitats (e.g., tree cover, plant species richness, and disturbance from roads) would be correlated with the variation in the spatial patterns of *A. pennsylvanica*.

@@

From each start site, we began walking along a path in a randomly selected direction (North, East, South, or West using a random number generator). We conducted a thorough visual search of the edges of the path for *A. pennsylvanica* funnel webs as the GPS receiver tracked our movement (Fig. S4). We walked along pre-existing sidewalks, roads, and trails because paths were a common feature of both habitats, along which we had previously found funnel-weaving spiders. We often found *A. pennsylvanica* in edge habitats in the study region, and this tendency to build webs at a habitat's edge is shared by other spider species [@cobbold_supp_2012]. Additionally, the thick undergrowth of the urban forest often restricted searches off the trails. To ensure equivalent search efforts, we took more time searching the edge of dense undergrowth in the urban forest.

@@

### Statistical analysis

@@

Since we found significant variation in distribution measures between habitats (see Results: Urban habitat characteristics), we wanted to explore environmental correlations with our findings. Specifically, we used the habitat subsets and the selected (non-correlated) environmental predictor variables to assess relationships between predictor variables and spatial pattern measures. In the global models, we included the predictors (without interactions) that remained after removal for collinearity with the response variables: web and spider abundance, nearest neighbor distance, and web height for the complete data and habitat subsets. We restricted these nearest neighbor analyses to first nearest neighbor distance (*i.e.,* excluded second nearest neighbor distance) because the first and second nearest neighbor distances both significantly differed between habitats, and including both would complicate the analysis. We excluded search distance as a response variable because the predictors were collected from the site of the focal web and may vary along the search path.

Several predictors (plant species richness, patch area, and road and highway disturbance) were highly right-skewed, so we used a natural log-transformation prior to model selection. We used Poisson generalized linear models for the number of webs and spiders and nearest neighbor distance and Poisson generalized linear mixed-effect models for web height with site as a random effect. From these global models, we performed Akaike Information Criterion (AIC) model selection with the *MuMIn* package in R [@MuMIn], which ranks models produced from every combination of the predictors in the global model. We reported the results of each top model, as well as a model average of models within 2 delta AICc points of the top model. We reported the estimates and standard error from the models without back-transformations for the log-transformed variables. We checked that the Variance of Inflation (VIF) stayed at or below a score of two for predictors in each top model using the *car* package in R [@car].

We also explored how land cover may affect web abundance, aggregation, and position using the National Land Cover Database and QGIS to record the category of the land cover class for the pixel that each site was located. We tested Poisson generalized linear models for each response variable with land cover as the only predictor. We used the *Anova()* function in the *car* package [@car] to get the Likelihood Ratio Chi-Squared values, then used the *glht()* function in the *multcomp* package [@multcomp] to perform pairwise comparisons among categories. We include these results in supplemental material (Fig. S10, Table S2) as the sample sizes were small (deciduous forest = 7, woody wetlands = 3, urban high-intensity = 4, urban medium-intensity = 6, urban low-intensity = 2).

@@

**Table 1** Landscape variables for the urban center (UNL City Campus), the urban forest (Wilderness Park), and Lincoln (Nebraska, USA) City Limits,  collected from the National Land Cover Database (2016 Tree Cover, 2019 Impervious Cover, 2019 Land Cover) using QGIS (v. 3.16.3-Hannover, ESRI 102704)

@@

All environmental predictors assessed within a 100-meter radius of the focal web varied by habitat (Table 2). Site tree cover was lower in the urban center than the urban forest, with five urban center sites measuring zero tree cover ($z_21$ = -14.50, *P* \< 0.001). The urban center sites had higher impervious cover, with all but four sites being zero in the urban forest ($z_21$ = 14.51, *P* \< 0.001). The urban center sites also had higher spectral radiance ($z_21$ = 3.08, *P* = 0.002) and artificial light at night radiance ($z_21$ = 22.46, *P* \< 0.001). Overall, plant species richness was lower in the urban center than the urban forest ($z_21$ = -7.67, *P* \< 0.001). Similarly, spiders used a lower diversity of plants as substrates in the urban center (mean ± se = 1.42 ± 0.15 species) than the urban forest (mean ± se = 2.7 ± 0.26 species, $z_21$ = -2.08, *P* = 0.037). Site patch area was smaller in the urban center ($z_21$ = -374.2, *P* \< 0.001). Road disturbance ($z_21$ = 39.19, *P* \< 0.001), highway disturbance ($z_21$ = 9.39, *P* \< 0.001), and road length ($z_21$ = -374.2, *P* \< 0.001) were higher for urban center sites than sites in the urban forest.

**Table 2** Mean and standard error of the collected environmental predictors by habitat. Percent tree and impervious cover, spectral radiance, artificial night sky radiance, and total road length were collected in a 100-meter buffer of each focal web, while plant species richness was recorded in a 10-meter buffer of each focal web. We used the nearest road or highway for disturbance measures. Patch area was defined as the area of continuous vegetation that includes the focal web

@@

**Fig. 3** Comparison of search distance (a), the number of webs (b) and spiders (c) in a 10-meter radius of the focal web, the distance from the focal web to the first and second (d) nearest neighbors, and web height (e) by habitat - urban center vs. urban forest. Grey jittered points represent the raw data, while black points and error bars represent the mean and 95% confidence intervals, respectively, from model predictions. Models were either Poisson Generalized Linear Models (a-c) or Poisson Generalized Linear Mixed Effect Models with site as a random factor (d-e). Asterisks indicate significance level (\*\*\* *P* \< 0.001; \*\* *P* \< 0.01; \* *P* \< 0.05). We searched shorter distances from the start site to the focal web (a), found more webs (b) and spiders (c) in a 10-meter radius of the focal web, measured shorter distances between the focal web and the two nearest neighbors (d), and recorded shorter web heights (e) in the urban center than the urban forest

@@

*Abundance* -- The top models for the number of webs and the number of spiders yielded similar results (Table 3, Fig. 4). For both measures of abundance, the urban forest subset retained no predictors (Table 3), while the urban center showed a decrease in abundance with increasing plant species richness, patch area, highway disturbance, and road length (Table 3, Fig. 4). The model average for the number of webs subset by urban center suggested similar results (Table S3). Plant species richness and highway disturbance both had a negative relationship with abundance measures for the overall data set (Table 3, Fig. S12). Interestingly, the top model of web abundance retained road disturbance and suggested a positive relationship overall (Table 3, Fig. S12).

@@

*Aggregation* -- Spiders were closer together across habitats when plant species richness was low and when road disturbance was high (Table 3, Fig. S12). In the urban forest subset, less tree cover, lower road and highway disturbance, and greater plant species richness were associated with nearness (Table 3, Fig. 5); however, this model was a poor fit ($R^2$ = 0.037). Urban center spiders were more aggregated where plant species richness, spectral radiance, and road length are high (Table 3, Fig. 5). The relationship between nearest neighbor distance and plant species richness is opposite when analyzed across habitats and within the urban subset. This is likely do to the differences in plant species richness and aggregation between habitats -- the urban center had low neighbor distance and plant diversity, while the urban forest had high neighbor distance and plant diversity. The model average for the nearest neighbor distance in the urban center corroborates the findings of the top model and adds that webs were clumped when patch area was larger (Table S3).

@@

For statistical analysis and figures of response variables by land cover class, see supplemental material (Table S2, Fig. S10). Following Likelihood Ratio tests, we find that all response variables significantly vary by land cover class (Table S2). There was no difference in search distance between urban medium-intensity, urban low-intensity, and deciduous forest. However, we searched significantly shorter in urban high-intensity and significantly further in woody wetlands (Fig. S10a). The number of webs and spiders only varied by habitat - urban low-, medium-, and high-intensity had more webs and spiders than deciduous forest or woody wetlands (Fig. S10b-c). For the urban intensities, web distance was shortest at low intensities and furthest at medium intensities with high urban intensity exhibiting web height between low and medium intensities (Fig. S10d). Nearest neighbor distance was shorter for woody wetlands than deciduous forest (Fig. S10d). Lastly, web height showed no difference between land cover class with pairwise comparisons (Fig. S10e).

@@

We also found that the webs of urban center spiders were built lower than webs in the urban forest, and this result was independent of the height of the substrate. One possible explanation for this pattern is that vibratory noise promotes lower web heights. Vibratory noise might, for example, be dampened on a plant nearer to the ground. We did find a trend that showed lower web heights overall at sites with higher road disturbance, but this idea requires further testing. Another potential explanation is that webs may vary in height based on the type of prey being ensnared. We have observed *Agelenopsis* spiders capture flying, hopping, and crawling insects. Arthropod community structure can be highly variable between urban habitats [@bang_faeth_2011], and urban centers and urban forests may support different types of prey items for *A. pennsylvanica*. Lower webs in the urban center may be ideal for more crawling prey, while higher webs in the urban forest may intercept more flying insects. Alternatively, webs in different habitats may experience different levels of disturbance. In the urban center, where lawns are mowed frequently and ornamental plants are regularly pruned, spiders could maintain lower webs to conserve energy from frequent rebuilding. Additionally, urban forests likely support (and act as a refuge for) many ground-dwelling vertebrates, whose movement along the forest floor could frequently damage lower webs. Greater tree coverage in the urban forest predicted higher webs, which may further suggest that small mammal movement could destroy low webs as areas with greater tree cover may provide protection and food resources for small mammals [@sarkar_2022]. Future research should assess the prey types available and small animal abundance and movement between these two habitats.

@@

We found that webs and spiders were more abundant (in the urban center) and more aggregated as plant species richness decreased. The urban center sites were covered in grass lawns and managed landscaping that usually had few plant species present but in high abundance. Sites with lower plant diversity may have few plant species that are highly preferred, leading to higher spider abundance. Research has shown mixed but often positive effects of plant diversity on spider abundance [@malumbres_et_al_2013; @koricheva_et_al_2000], suggesting that plant community composition may play a vital, yet context-specific role in spider spatial patterns. To better understand the role that plant species play in abundance would require comparing plant communities between patches used by spiders versus those devoid of spiders. We speculate that greater aggregation on more diverse plant communities may suggest that *A. pennsylvanica* prefers a certain plant or type of plant, especially in the urban center where we observed similar plants clumped in space.

@@

Lower road and highway disturbance was associated with larger spider aggregations in the urban forest but had no effect on urban center spiders. Perhaps noise negatively affects urban forest spiders -- encouraging spiders to inhabit areas further from noise sources and reducing space between neighbors. Many spiders are thought to be plastic in the structure and tuning of their webs [@mortimer_2016; @blamires_2010], and urban exploiters are often fast to adjust to changing conditions [@kralj_2017]. As such, spiders reared in or descended from spiders in the urban center might be better able to tune their webs to decrease vibrations, although this hypothesis requires further testing. It will also be important to differentiate whether these differences in behavior between environments are based on developmental exposure, behavioral plasticity, or genetic differentiation.

@@

Lastly, spiders in the urban center were more aggregated at sites with higher spectral radiance (a measure of surface temperature). Urban centers tend to be warmer on average compared to surrounding areas due to heat-trapping via pavement in a phenomenon called the Urban Heat Island Effect [@johnson_et_al_2020]. Aggregation formation in warmer sites could be the result of preference for or acclimation to warmer areas [@meineke_et_al_2013], or perhaps aggregation occurs at limited thermal refuges [@riechert_1974b].

@@

We have shown that human-caused environmental changes correlate with spider spatial patterns, but there are likely many other forms of human disturbance than those tested here. Our study shows that a solitary, generalist predator (*A. pennsylvanica*) is more abundant and aggregated with lower web height in an urban center compared to an urban forest -- habitats that vary significantly in environmental conditions. We did not measure community-level interactions among organisms, such as the effects of prey abundance; however, correlations between spatial patterns and environmental conditions indirectly indicate that such interspecific interactions may be important influencers of spider aggregations. Noise may also greatly influence spider spatial patterns. Our variables that estimate the potential for vibratory noise negatively affected spider abundance while promoting lower web heights, potentially to reduce interference from vibratory noise. Noise variables also had opposite effects on aggregative behavior of spiders in the different habitats. These differences may indicate variability in tolerance to noise, but more research is necessary to test this hypothesis.
